---
title: "National and state excess deaths analysis"
author: "Andrew Ba Tran"
date: "6/30/2020"
output:
  html_document:
    code_folding: hide
---

```{r prep, warning=F, message=F}

packages <- c("tidyverse", "lubridate", "writexl", "knitr",
              "MMWRweek")

if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())), repos = "http://cran.us.r-project.org")  
}

options(knitr.kable.NA = '')

library(tidyverse)
library(lubridate)
library(writexl)
library(knitr)
library(MMWRweek)

nat <- read_csv("data/outputs/national_and_state_summary.csv") %>% 
  filter(state=="US.agg") %>% 
  rename(covid19.nchs=covid.death.hybrid)

dates <- MMWRweek(nat$week_start_date)

nat$year <- dates$MMWRyear
nat$day <- dates$MMWRday
nat$week <- dates$MMWRweek
```

```{r national_weekly_chart, warning=F, message=F, fig.width=9, fig.height=4}
weeks_in <- 1:24
weeks_march <- 10:24

natlsum_ad <- nat %>%
  filter(year==2020) %>% 
  filter(week %in% weeks_in)

national_weekly <- natlsum_ad %>% 
  select(week, week_end_date, year,
                     all_cause_deaths, expected_all_cause=baseline_all_cause,
                     expected_all_cause_upper=baseline_all_cause_upper, 
                     expected_all_cause_lower=baseline_all_cause_lower,
                     excess_all_cause_deaths, covid19.nchs)

natlsum_ad %>% 
  ggplot(aes(week_end_date,all_cause_deaths)) +
  geom_ribbon(aes(ymin=baseline_all_cause_lower, ymax=baseline_all_cause_upper), fill="gray70", alpha=.5) +
  geom_ribbon(aes(ymin=baseline_all_cause, ymax=all_cause_deaths), fill="sienna1", alpha=.5) +
  geom_ribbon(aes(ymin=all_cause_deaths-covid19.nchs, ymax=all_cause_deaths), fill="tomato3", alpha=.5) +
  geom_line(data=natlsum_ad, aes(x=week_end_date, y=all_cause_deaths),color="sienna1", alpha=.4) +
  geom_line(color="black", size=1) +
  theme_minimal() +
  labs(title="U.S. Deaths", y="Weekly deaths", x="") 

```

```{r states, warning=F, message=F}

state_lag <- read_csv("data/outputs/NobBs.complete.csv")

state.abb <- c(state.abb, "DC", "NYC")
state.name <- c(state.name, "District of Columbia", "New York City")
state_names <- data.frame(state.abb, state.name)

state_lag_long <- state_lag %>%
  filter(week==4) %>% 
  pivot_longer(cols=3:ncol(state_lag),
               names_to="state",
               values_to="lag") %>% 
  mutate(quantile_rank=ntile(lag, 3)) %>% 
  select(-X1, -week) %>% 
  left_join(state_names, by=c("state"="state.name")) %>% 
  select(-state) %>% 
  rename(state=state.abb)

states <- read_csv("data/outputs/national_and_state_summary.csv") %>% 
  filter(state!="US" & state!="US.agg") %>% 
  rename(covid19.nchs=covid.death.hybrid) %>% 
  ungroup() %>% 
  left_join(state_lag_long)

orders <- read_csv("data/archive/stay_at_home_orders.csv") %>%   
  left_join(state_names, by=c("state"="state.name")) %>% 
  select(-state, state=state.abb)

states <- left_join(states, orders)

states <- states %>% 
  mutate(restriction_type=gsub("_variance", "", restrictions))
states$restriction_type = ifelse(states$state=="NYC", "major", states$restriction_type)
```

```{r week15, warning=F, message=F}
states_23 <- states %>% 
  filter(year==2020) %>% 
  filter(week==23) %>% 
  mutate(flag = case_when(
    all_cause_deaths < baseline_all_cause_lower ~ "Lower than baseline range",
    all_cause_deaths > baseline_all_cause_upper ~ "Higher than baseline range",
    TRUE ~ "Within range"
  )) %>% 
  select(state, flag23=flag)

states_24 <- states %>% 
  filter(year==2020) %>% 
  filter(week==24) %>% 
  mutate(flag = case_when(
    all_cause_deaths < baseline_all_cause_lower ~ "Lower than baseline range",
    all_cause_deaths > baseline_all_cause_upper ~ "Higher than baseline range",
    TRUE ~ "Within range"
  )) %>% 
  select(state, flag24=flag)
```

```{r beautiful_states, warning=F, message=F}
march <- states %>% 
  filter(year==2020) %>% 
  filter(week %in% weeks_march) %>% 
  group_by(state) %>%
  summarize(all_cause_deaths=sum(all_cause_deaths, na.rm=T),
            expected=sum(baseline_all_cause, na.rm=T),
            excess_deaths=sum(excess_all_cause_deaths, na.rm=T),
            covid_deaths=sum(covid19.nchs, na.rm=T),
            x_excess_covid=round(excess_deaths/covid_deaths,1),
            excess_minus_covid=excess_deaths-covid_deaths
            ) %>% 
  mutate(percent_covid_in_excess=round(covid_deaths/excess_deaths*100)) %>% 
  ungroup() %>% 
  left_join(states_23) %>% 
  left_join(states_24) 


march$percent_covid_in_excess <- ifelse(march$percent_covid_in_excess >100, NA, march$percent_covid_in_excess)
march$percent_covid_in_excess <- ifelse(march$percent_covid_in_excess <0, NA, march$percent_covid_in_excess)

march <- unique(march)
march <- march %>% 
  mutate(percent_more=round((all_cause_deaths-expected)/expected*100,2))
march <- march %>% 
  left_join(state_lag_long) %>% 
  left_join(orders)  %>% 
  mutate(restriction_type=gsub("_variance", "", restrictions))

higher <- march %>% 
  filter(flag24=="Higher than baseline range") %>% 
  pull(state)

within <- march %>% 
  filter(flag24=="Within range") %>% 
  pull(state)

lower <- march %>% 
  filter(flag24=="Lower than baseline range") %>% 
  pull(state)
march$restriction_type = ifelse(march$state=="NYC", "major", march$restriction_type)


major <- march %>% 
  filter(restriction_type=="major") %>% 
  pull(state)

minor <- march %>% 
  filter(restriction_type=="minor") %>% 
  pull(state)

moderate <- march %>% 
  filter(restriction_type=="moderate") %>% 
  pull(state)


none <- march %>% 
  filter(restriction_type=="none") %>% 
  pull(state)




```

```{r pretty_table2, warning=F, message=F}
nat_solo <- national_weekly %>% 
  filter(year==2020) %>% 
  filter(week %in% weeks_march) %>% 
  #filter(week %in% 10:14) %>% 
  ungroup() %>% 
  summarize(all_deaths=sum(all_cause_deaths),
            excess_deaths=sum(excess_all_cause_deaths),
            covid19.nchs=sum(covid19.nchs)) %>% 
  mutate(percent_covid_in_excess=round(covid19.nchs/excess_deaths*100)) %>% 
  mutate(Place="U.S.") %>% 
  select(Place, all_deaths, excess_deaths, covid19.nchs, percent_covid_in_excess)

flagged_states <- march %>% 
# change flag here maybe
    filter(flag23=="Higher than baseline range" & flag24=="Higher than baseline range") %>% 
  filter(excess_deaths >0) %>% 
  filter(x_excess_covid >1) %>% 
  select(Place=state, all_deaths=all_cause_deaths,
         excess_deaths=excess_deaths, covid19.nchs=covid_deaths,
         percent_covid_in_excess) 

nice_table <- rbind(nat_solo, flagged_states) %>% 
  mutate(excess_minus=excess_deaths-covid19.nchs) %>% 
  rename(`All deaths`=all_deaths,
         `Excess deaths`=excess_deaths,
         `Covid-19 deaths`=covid19.nchs,
         `Covid-19 % of excess` = percent_covid_in_excess,
         `Excess deaths minus covid-19`=excess_minus)

other_states <- march %>% 
  filter(flag23!="Higher than baseline range" | flag24!="Higher than baseline range")
other_states$percent_covid_in_excess <- ifelse(other_states$percent_covid_in_excess > 100, NA, other_states$percent_covid_in_excess)
other_states$percent_covid_in_excess <- ifelse(other_states$percent_covid_in_excess == 0, NA, other_states$percent_covid_in_excess)
other_states$excess_minus_covid <- ifelse(other_states$excess_minus_covid<0, NA, other_states$excess_minus_covid)

other_states <- other_states %>%   
  select(Place=state, `All deaths`=all_cause_deaths,
         `Excess deaths`=excess_deaths, `Covid-19 deaths`=covid_deaths,
         `Covid-19 % of excess`=percent_covid_in_excess, `Week 23`=flag23, `Week 24`=flag24) 
```

There were around 135,500 excess deaths between March and mid June, about 34 percent more than were publicly attributed to covid-19 at the time.

This updated model adjusts not just for seasonality, population, flu, but also accounts for delay in reporting from individual states. 

More recent data suggests that the gap between unattributed and attributed covid-19 deaths is shrinking. 

## Deaths through June 13, 2020

```{r ustable}
kable(nice_table %>% filter(Place=="U.S."))
```


## States 

```{r state_charts_major, warning=F, message=F, fig.height=16, fig.width=9}
states %>% 
  filter(year==2020) %>% 
  filter(week %in% weeks_in) %>% 
  ggplot(aes(week_end_date,all_cause_deaths)) +
  geom_ribbon(aes(ymin=baseline_all_cause_lower, ymax=baseline_all_cause_upper), fill="gray70", alpha=.5) +
  geom_ribbon(aes(ymin=baseline_all_cause, ymax=all_cause_deaths), fill="sienna1", alpha=.5) +
  geom_ribbon(aes(ymin=all_cause_deaths-covid19.nchs, ymax=all_cause_deaths), fill="tomato", alpha=.5) +
  geom_line(aes(x=week_end_date, y=all_cause_deaths),color="sienna1", alpha=.4) +
  geom_line(color="black", size=.5) +
  facet_wrap(~state, scales="free_y", ncol=5) +
  theme_minimal()#, 
       #subtitle="Exceeds the expected range and with the least delays in reporting")

```

```{r state_table}


march$percent_covid_in_excess <- ifelse(march$percent_covid_in_excess > 100, NA, march$percent_covid_in_excess)
march$percent_covid_in_excess <- ifelse(march$percent_covid_in_excess == 0, NA, march$percent_covid_in_excess)
march$excess_minus_covid <- ifelse(march$excess_minus_covid<0, NA, march$excess_minus_covid)

march <- march %>%   
  mutate(lag=case_when(
    quantile_rank==1 ~ "Significant reporting lag",
    quantile_rank==2 ~ "Some reporting lag",
    quantile_rank==3 ~ "Least reporting lag"
  )) %>% 
  select(Place=state, `All deaths`=all_cause_deaths,
         `Excess deaths`=excess_deaths, `Covid-19 deaths`=covid_deaths,
#         `Covid-19 % of excess`=percent_covid_in_excess, `Week 23`=flag23, `Week 24`=flag24, lag, order=restriction_type) 
         `Covid-19 % of excess`=percent_covid_in_excess, `Week 23`=flag23, `Week 24`=flag24, lag) 

kable(march %>% arrange(`Covid-19 % of excess`))
```

```{r export, warning=F, message=F}

states <- states %>% 
  filter(year==2020) %>% 
  mutate(adjusted_expected=all_cause_deaths-baseline_all_cause,
         adjusted_expected_minus_covid=adjusted_expected-covid19.nchs)

state_lag_narrow <- state_lag_long %>% 
  mutate(state=as.character(state)) %>% 
  select(Place=state, lag_index=lag) %>% 
  ungroup()

march <- left_join(march, state_lag_narrow)

write_csv(march, "data/outputs/states_summary_table.csv", na="")
write_csv(national_weekly, "data/outputs/national_weekly_table.csv", na="")
write_csv(states, "data/outputs/states_weekly_table.csv", na="")
write_xlsx(list(`beautiful table`= nice_table,
                `states summary table` = march,
                `national weekly` = national_weekly,
                `states weekly`=states), 
           "data/outputs/beautiful_table.xlsx")


```