---
title: "Fluview vs NCHS All Deaths"
author: "Andrew Ba Tran"
date: "7/1/2020"
output: html_document
---
```{r setup, warning=F, message=F}
library(tidyverse)
library(janitor)
library(lubridate)
library(knitr)
```

# State gap

```{r load_data, warning=F, message=F}
cdc1 <- read_csv("https://data.cdc.gov/api/views/r8kw-7aab/rows.csv?accessType=DOWNLOAD")
cdc1 <- clean_names(cdc1) %>% 
  mutate(end_week=mdy(end_week)) %>% 
  group_by(state) %>%
  arrange(end_week) %>% 
  mutate(week=row_number()+4)

fluview <- read_csv("data/weekly_fluview/State_Custom_data25.csv")

fluview <- clean_names(fluview) %>% 
  filter(season=="2019-20") %>% 
  select(week, state=sub_area, total_deaths2=total_deaths) %>% 
  mutate(total_deaths2=gsub(",", "", total_deaths2)) %>% 
  mutate(total_deaths2=as.numeric(total_deaths2))

cdc3 <- left_join(cdc1, fluview)

us1 <- cdc3 %>% 
    filter(week<=23) %>% 

  mutate(check_all=case_when(
    total_deaths >  total_deaths2 ~ "More",
    total_deaths  ==  total_deaths2 ~ "Equal",
    total_deaths  <  total_deaths2 ~ "Less"
  )) %>% 
  select(start_week, end_week, state, nchs=total_deaths, 
         fluview=total_deaths2) %>% 
  filter(state!="United States") %>% 
  filter(state!="Puerto Rico") %>% 
  pivot_longer(cols=4:5, names_to="type", values_to="deaths")
```

```{r chart1, warning=F, message=F, fig.height=12, fig.width=9}
ggplot(us1, aes(x=end_week, y=deaths, color=type)) +
  geom_line() +
  facet_wrap(~state, ncol=4, scales="free_y")
```

```{r summary, warning=F, message=F}
gap_state <- cdc3 %>% 
      filter(week<=23) %>% 

  group_by(state) %>% 
  summarize(nchs=sum(total_deaths, na.rm=T),
            fluview=sum(total_deaths2, na.rm=T)) %>% 
    filter(fluview!=0) %>% 
  mutate(difference=nchs-fluview)
kable(gap_state, format.args = list(big.mark = ","))
```

## United States

```{r combined, warning=F, message=F}
gap_state_summary <-
  gap_state %>% 
  summarize(nchs=sum(nchs),
            fluview=sum(fluview))

kable(gap_state_summary, format.args = list(big.mark = ","))
```

Presumably, NCHS and Fluview data should even out, right? 

The NCHS data counts people who died in the state regardless of residence.

The Fluview data counts people who died but were residents. 

So what happens to those who weren't residents? They're added to the Fluview data to their states of origin?

So why is there a gap of `r prettyNum(gap_state_summary$nchs - gap_state_summary$fluview, big.mark=",")` deaths? 

Who has fallen between the gaps? Are these undocumented? Internationals?

